---
title: "Comparing Network Indicators and Graphon Distances for Structural Anomaly Detection"
subtitle: "A Reproducible Workflow for Directed Weighted Longitudinal Networks"
author: "Zsolt T. Kosztyán and Kornél Dénes"
date: ""
output:
  html_document:
    toc: true
    toc_depth: 2
  officedown::rdocx_document:
    reference_docx: output/MethodsX-reference-clean.docx
    toc: false
    number_sections: false
params:
  build_figures: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 300,
  eval = isTRUE(params$build_figures)
)

required_packages <- c(
  "ggplot2", "dplyr", "tidyr", "scales",
  "officedown", "officer", "flextable"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Missing R package(s): ", paste(missing_packages, collapse = ", "))
}
suppressPackageStartupMessages(library(dplyr))

graphical_abstract_script <- file.path(
  "scripts", "production", "06_graphical_abstract.R"
)
if (!file.exists(graphical_abstract_script)) {
  stop("Graphical-abstract source missing: ", graphical_abstract_script)
}
source(graphical_abstract_script, local = TRUE)

figure_dir <- file.path("output", "figures")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

figure_inputs <- c(
  simulation_kernels = file.path(
    "data", "derived", "simulation", "SIMULATION_kernel_design.csv"
  ),
  simulation_power = file.path(
    "data", "derived", "simulation", "SIMULATION_detection_summary.csv"
  ),
  empirical_evidence = file.path(
    "data", "derived", "empirical", "EMPIRICAL_transition_evidence.csv"
  ),
  estimator_agreement = file.path(
    "data", "derived", "empirical", "EMPIRICAL_estimator_agreement.csv"
  ),
  resolution_agreement = file.path(
    "data", "derived", "empirical", "EMPIRICAL_resolution_agreement.csv"
  ),
  role_flow = file.path(
    "data", "derived", "empirical", "EMPIRICAL_role_flow_matrix_2018.csv"
  ),
  refinement = file.path(
    "data", "derived", "empirical", "EMPIRICAL_multiresolution_refinement.csv"
  )
)
if (isTRUE(params$build_figures)) {
  missing_inputs <- figure_inputs[!file.exists(figure_inputs)]
  if (length(missing_inputs)) {
    stop(
      "Figure input(s) missing. Run source('run_all.R') first: ",
      paste(missing_inputs, collapse = ", ")
    )
  }
}

mx_colors <- c(
  blue = "#2F6B8A",
  gold = "#C9912B",
  orange = "#D56A2D",
  olive = "#6F7D3C",
  pink = "#B85C7A",
  ink = "#25313A",
  mid_grey = "#8B959C",
  light_grey = "#E7EBEE"
)

theme_methodsx <- function(base_size = 10) {
  ggplot2::theme_minimal(base_size = base_size, base_family = "sans") +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggplot2::element_text(
        color = mx_colors["ink"], face = "bold", size = base_size + 2
      ),
      plot.subtitle = ggplot2::element_text(color = "#4F5B62", size = base_size),
      plot.caption = ggplot2::element_text(color = "#657078", size = base_size - 2),
      axis.title = ggplot2::element_text(color = mx_colors["ink"]),
      axis.text = ggplot2::element_text(color = mx_colors["ink"]),
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major = ggplot2::element_line(color = mx_colors["light_grey"], size = 0.3),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(face = "bold"),
      strip.text = ggplot2::element_text(face = "bold", color = mx_colors["ink"])
    )
}

save_methodsx_figure <- function(plot, stem, width, height) {
  ggplot2::ggsave(
    file.path(figure_dir, paste0(stem, ".pdf")),
    plot = plot, width = width, height = height, units = "in",
    device = grDevices::pdf
  )
  ggplot2::ggsave(
    file.path(figure_dir, paste0(stem, ".png")),
    plot = plot, width = width, height = height, units = "in",
    dpi = 600, bg = "white"
  )
  invisible(plot)
}

channel_labels <- c(
  indicator_total_weight = "Total weight",
  indicator_edge_density = "Edge density",
  indicator_weighted_reciprocity = "Weighted reciprocity",
  indicator_in_strength_gini = "In-strength Gini",
  indicator_out_strength_gini = "Out-strength Gini",
  indicator_effective_destinations = "Effective destinations",
  indicator_top10_destination_share = "Top-10 destination share",
  indicator_binary_transitivity = "Binary transitivity",
  graphon_frobenius_raw = "Frobenius distance: raw scale",
  graphon_cut_raw = "Cut-norm lower bound: raw scale",
  graphon_wasserstein_raw = "Wasserstein distance: raw scale",
  graphon_frobenius_shape = "Frobenius distance: unit-mean shape",
  graphon_cut_shape = "Cut-norm lower bound: unit-mean shape",
  graphon_wasserstein_shape = "Wasserstein distance: unit-mean shape",
  singular_spectrum_raw = "Raw singular-spectrum distance",
  singular_spectrum_shape = "Shape singular-spectrum distance",
  estimated_role_instability = "Estimated role instability"
)
```

# Article information

## Article title

**Comparing Network Indicators and Graphon Distances for Structural Anomaly Detection**

*A Reproducible Workflow for Directed Weighted Longitudinal Networks*

## Authors

Zsolt T. Kosztyán^a,b,c^; Kornél Dénes^d^

## Affiliations

a. Business Research Institute for Data-Driven Growth and Economics (BRIDGE), Faculty of Business and Economics, University of Pannonia, Egyetem Street 10, 8200 Veszprém, Hungary

b. Department of Quantitative Methods, Institute of Management, Faculty of Business and Economics, University of Pannonia, Egyetem Street 10, 8200 Veszprém, Hungary

c. Institute of Advanced Studies, Kőszeg (iASK), Chernel Street 14, 9730 Kőszeg, Hungary

d. Faculty of Law, Institute of Political Science, Eötvös Loránd University, Tóth Kálmán Street 10, 1097 Budapest, Hungary

## Corresponding author's email address and Twitter handle

kosztyan.zsolt@gtk.uni-pannon.hu; Twitter/X: not applicable

## Keywords

longitudinal networks; anomaly detection; network indicators; graphon distances;
directed weighted networks; block-kernel estimation; comparative method evaluation

## Related research article

Kosztyán, Z. T., Hosznyák, A., Király, T., Katona, A. I., Dénes, K., and
Hornák, G. *Tracking Structural Evolution in Higher Education Application
Networks with Graphon Distances: Hungary, 2006--2024*. Companion research
manuscript; publication details will be inserted when available.

# Abstract

Structural anomaly detection in longitudinal networks is difficult because
year-specific network indicators and matrix-level distances answer different
questions. This article presents a reproducible workflow that compares and
integrates two complementary evidence families: conventional indicators of activity,
connectivity, reciprocity, and concentration, and graphon-inspired distances
between node-aligned fitted directed weighted kernels. An anomaly-scoring layer
converts heterogeneous transition measures to comparable within-channel
evidence, groups correlated channels into structural perspectives, and reports
their consensus without replacing the component results. Raw-weight and
unit-mean comparisons distinguish changes in magnitude from redistribution of
structural shape. Known-truth simulations calibrate channel-specific null
thresholds and demonstrate complementarity: conventional indicators are highly
sensitive and directly interpretable for global activity, density, and
concentration changes, whereas fitted-kernel and spectral distances add
sensitivity to coordinated relational changes. The maximum null false-positive
rate is 5%; at the medium effect, the raw cut-norm lower-bound estimate detects
86.0% of directional shifts, 61.5% of local block shocks, and 59.5% of role
reassignments. In 19 annual Hungarian higher-education application networks,
2017--2018 has the strongest consensus, with five of six perspectives at or
above the 75th percentile. The method treats graphon distances as a complement
to, not a substitute for, conventional longitudinal network analysis.

# Graphical abstract

```{r graphical-abstract, fig.width=6.8, fig.height=3.40, fig.alt="Conceptual workflow in which annual networks feed two complementary evidence lanes: indicator time series with a highlighted anomaly and graphon-inspired distances illustrated by a fitted surface. A calibration target leads to agreement or disagreement as interpretable outcomes."}
p_graphical_abstract <- build_methodsx_graphical_abstract(mx_colors)
save_methodsx_figure(
  p_graphical_abstract, "FIG00_graphical_abstract", 12, 6
)
p_graphical_abstract
```

The graphical abstract presents the study as a comparison of two equally
important evidence families rather than as a graphon-only pipeline. Annual
network indicators preserve direct substantive interpretation; graphon-inspired
distances retain coordinated relational information. Their convergence supports
a broad structural-anomaly claim, while disagreement identifies the type and
scale of change that requires interpretation.

# Specifications table

```{r specifications-table, results='asis'}
specifications <- data.frame(
  field = c(
    "Subject area",
    "More specific subject area",
    "Name of your method",
    "Name and reference of original method",
    "Resource availability"
  ),
  entry = c(
    "Computer Science",
    "Network science; temporal and dynamic network analysis",
    "Network-indicator and graphon-distance comparison workflow",
    paste(
      "Annual network indicators, directed weighted block-kernel estimation,",
      "node-aligned distances, and anomaly-score consensus; see Donnat and",
      "Holmes [1], Lovász [2], and the companion research article listed above"
    ),
    paste(
      "R source code, aggregate annual networks, intermediate analytical",
      "outputs, validation tables, and figure-generation code are supplied",
      "in the accompanying reproducibility package"
    )
  ),
  check.names = FALSE
)
spec_ft <- flextable::flextable(specifications)
spec_ft <- flextable::delete_part(spec_ft, part = "header")
spec_ft <- flextable::bold(spec_ft, j = 1, part = "body", bold = TRUE)
spec_ft <- flextable::width(spec_ft, j = 1, width = 2.15)
spec_ft <- flextable::width(spec_ft, j = 2, width = 4.10)
table_border <- officer::fp_border(color = "000000", width = 0.75)
spec_ft <- flextable::border_remove(spec_ft)
spec_ft <- flextable::border_outer(
  spec_ft, border = table_border, part = "all"
)
spec_ft <- flextable::border_inner_h(
  spec_ft, border = table_border, part = "body"
)
spec_ft <- flextable::border_inner_v(
  spec_ft, border = table_border, part = "body"
)
spec_ft <- flextable::padding(spec_ft, padding = 5, part = "all")
spec_ft <- flextable::valign(spec_ft, valign = "center", part = "all")
spec_ft <- flextable::font(spec_ft, fontname = "Calibri", part = "all")
spec_ft <- flextable::fontsize(spec_ft, size = 10.5, part = "all")
spec_ft
```

# Background

Longitudinal network analysis commonly begins with annual indicators such as
total weight, density, reciprocity, transitivity, or concentration. These
measures are substantively interpretable and remain the first evidence family
in the proposed workflow, but each compresses an entire relational system to a
scalar. A transition can therefore rearrange coordinated origin--destination
patterns while producing only a modest change in an aggregate indicator. At
the other extreme, relying on one matrix distance can make the result difficult
to interpret and can miss the form of change that matters. Reviews of graph
comparison accordingly emphasize that the meaning of a distance depends on the
network representation and the structural feature being compared [1,3].

This method combines rather than ranks these approaches. Conventional network
indicators describe interpretable annual trajectories of activity,
connectivity, reciprocity, and concentration. Graphon-inspired fitted-kernel
distances form a second evidence family that retains node-aligned relational
structure at a selected resolution. The raw weighted representation asks
whether the intensity surface changed; unit-mean normalization asks whether
weights were redistributed after removing the average level. Frobenius
distance accumulates cellwise deviations, a cut-norm search targets a coherent
rectangular origin--destination subset, Wasserstein distance compares marginal
fitted-intensity distributions, and singular values summarize dominant
directed modes of the original adjacency matrices.

An anomaly-scoring layer links the two evidence families. Each empirical
transition is ranked relative to the other observed transitions within its own
channel, and correlated channels are summarized into six perspectives before a
consensus profile is calculated. Known-truth simulations separately calibrate
channel-specific null thresholds. Agreement across indicator and graphon-based
perspectives supports a broader structural claim; disagreement is retained as
diagnostic information about the kind of change observed. Graphon distances
therefore complement rather than replace conventional network analysis.

The workflow is designed for node-aligned directed weighted snapshots. It does
not minimize over node relabelings, and its fitted non-negative intensity
kernels are block-kernel analogues rather than probability-valued graphons.
These distinctions are important: graphon theory motivates the representation
[2,4], but the empirical objects and the reported discrepancies retain node
identity and raw weight scale.

The practical contribution is a guide to method selection and claim strength.
Researchers retain familiar annual network indicators, add graphon-inspired
distances where relational detail is needed, and use anomaly profiles to decide
whether a transition reflects scale, broad matrix change, localized coordinated
change, redistribution, dominant spectral modes, or role reorganization.
Simulation calibration and specification sensitivity are part of the method,
not optional post-hoc diagnostics.

The comparison is organized around three research questions. **RQ1:** How do
conventional annual network indicators and graphon-inspired distances differ in
their sensitivity to known forms of longitudinal network change? **RQ2:** When
do the two evidence families converge or diverge in the empirical ranking of
annual transitions? **RQ3:** How can null calibration, estimator and resolution
sensitivity, and role-flow interpretation translate those patterns into
reproducible claims? These questions determine the sequence of the validation
and empirical sections and provide the conceptual structure of the graphical
abstract.

The central methodological question is therefore not which single metric is
best, but how interpretable network indicators and relational distance measures
can be combined without concealing their different sensitivities. The main
contribution is the integration protocol: preserve the annual indicator
trajectories, add graphon-inspired distance channels, calibrate their detector
behaviour under known perturbations, reduce correlated channels to transparent
perspectives, and interpret both consensus and disagreement. It is a workflow
contribution rather than a claim that one newly proposed distance dominates all
existing alternatives.

# Method details

## Design principles and workflow

The method requires a sequence of non-negative directed weighted adjacency
matrices $A_t\in\mathbb{R}_{\geq0}^{n\times n}$ with the same ordered node set
in every snapshot. Self-loops may be retained as substantive weights, but
density and binary transitivity are calculated after removing the diagonal.
All preprocessing decisions, including the node universe, time interval,
weight definition, and missing-node rule, must be fixed before transition
scores are calculated.

Structural change is not a single observable quantity. A transition may alter
overall activity, connectivity, concentration, directionality, mesoscale
organization, or the roles occupied by individual nodes. Figure 1 presents the
method as two evidence families joined by a common anomaly-scoring and decision
layer. The resulting profile is interpreted only after known-truth calibration
and robustness assessment.

```{r figure-workflow, fig.width=6.8, fig.height=5.95, fig.cap="Reproducible workflow for detecting, validating, and interpreting structural anomalies in node-aligned directed weighted longitudinal networks."}
workflow_nodes <- data.frame(
  id = 1:7,
  x = c(2, 1, 3, 2, 1, 3, 2),
  y = c(6, 5, 5, 4, 3, 3, 2),
  label = c(
    "1  Input integrity\nnode alignment + weights",
    "2A  Conventional\nnetwork indicators",
    "2B  Graphon-inspired\nkernel distances",
    "3  Channel-specific\nanomaly scores",
    "4A  Known-truth null\ncalibration",
    "4B  Estimator + resolution\nrobustness",
    "5  Perspective consensus +\nstructural interpretation"
  ),
  group = c("input", "branch", "branch", "evidence", "validation", "validation", "interpretation")
)
workflow_edges <- data.frame(
  x = c(2, 2, 1, 3, 2, 2, 1, 3),
  y = c(5.72, 5.72, 4.72, 4.72, 3.72, 3.72, 2.72, 2.72),
  xend = c(1, 3, 2, 2, 1, 3, 2, 2),
  yend = c(5.30, 5.30, 4.30, 4.30, 3.30, 3.30, 2.30, 2.30)
)
workflow_nodes$fill <- c(
  input = "#E8F1F5", branch = "#F2F5F7", evidence = "#E8F1F5",
  validation = "#FBF1DE", interpretation = "#EDF0E3"
)[workflow_nodes$group]

p_workflow <- ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = workflow_edges,
    ggplot2::aes(x = x, y = y, xend = xend, yend = yend),
    size = 0.55, color = mx_colors["mid_grey"],
    arrow = grid::arrow(length = grid::unit(0.13, "inches"), type = "closed")
  ) +
  ggplot2::geom_tile(
    data = workflow_nodes,
    ggplot2::aes(x = x, y = y, fill = fill),
    width = 1.55, height = 0.62, color = mx_colors["ink"], size = 0.35
  ) +
  ggplot2::geom_text(
    data = workflow_nodes,
    ggplot2::aes(x = x, y = y, label = label),
    size = 3.4, lineheight = 0.95, color = mx_colors["ink"]
  ) +
  ggplot2::scale_fill_identity() +
  ggplot2::coord_cartesian(xlim = c(0.05, 3.95), ylim = c(1.55, 6.45), clip = "off") +
  ggplot2::labs(
    title = "Integrated structural-anomaly workflow",
    subtitle = "Two evidence families are integrated by anomaly scoring, validation, and consensus"
  ) +
  ggplot2::theme_void(base_family = "sans") +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", color = mx_colors["ink"], size = 13),
    plot.subtitle = ggplot2::element_text(color = "#4F5B62", size = 10),
    plot.title.position = "plot",
    plot.margin = ggplot2::margin(10, 12, 10, 12)
  )
save_methodsx_figure(p_workflow, "FIG01_workflow", 8, 7)
p_workflow
```

Conventional indicators and graphon-inspired distances provide parallel
transition evidence. The anomaly layer places their heterogeneous scores on
comparable within-channel scales, while simulations establish which mechanisms
each channel can detect and specification checks reveal which rankings remain
stable. Consensus summarizes convergence without erasing disagreement. Role
and flow analysis is used last, to interpret rather than define an anomaly.

## Step 1. Construct and verify aligned snapshots

For each time point, convert the edge list to an $n\times n$ matrix in the same
node order. The included empirical application contains 19 annual snapshots
from 2006 through 2024 and 175 nodes in every year. The pipeline verifies the
year sequence, matrix dimensions, finite non-negative weights, and two
regression anchors: total weight 358,367 in 2016 and 384,035 in 2019. These
anchors detect accidental use of the superseded 2016 input or an incorrectly
copied annual table.

## Step 2. Calculate interpretable annual network indicators

Eight snapshot descriptors provide substantive anchors: total weight,
off-diagonal edge density, weighted reciprocity, in- and out-strength Gini
coefficients, the entropy-based effective number of destinations, the share of
weight received by the ten largest destinations, and binary transitivity.
Transition scores are absolute consecutive-year changes; total weight and
effective destinations are compared on a $\log(1+x)$ scale. These descriptors
are not treated as interchangeable tests. Each describes a specific structural
dimension that helps explain a distance peak.

## Step 3. Fit directed weighted block kernels

At each fixed resolution $K$, nodes are clustered from standardized outgoing
strength, incoming strength, outgoing degree, and incoming degree. For the
reference block estimator, the fitted block-pair intensity is the mean of the
corresponding entries of $A_t$. If $z_t(i)$ is the block assigned to node $i$
and $W_t$ is the fitted $K\times K$ block matrix, the node-aligned fitted kernel
is [8]

$$
M_t(i,j)=W_t[z_t(i),z_t(j)].
$$

The reference temporal analysis holds $K=13$ constant. It is a common
high-resolution comparison scale, not an estimate of a unique true number of
roles. Robustness is evaluated with block, smoothed, and spectral estimators at
$K=13$, and with the block estimator at $K\in\{2,4,8,13\}$.

The snapshot-specific adaptive-$K$ heuristic is diagnostic only. For
$n\leq200$, its corrected implementation uses deterministic dense `svd()`,
removes singular values no larger than $10^{-8}$ of the leading singular value,
and then applies the relative-drop rule. This prevents an iterative singular-
value solver's numerical noise floor from forcing rank-deficient matrices to
the $\lfloor\sqrt{n}\rfloor$ ceiling. In the present data the corrected
diagnostic gives $K=2$ for most years and $K=13$ for 2009--2011; it does not
change any fixed-$K=13$ distance reported here.

## Step 4. Compare raw intensity and normalized shape

For each adjacent pair, compute every distance twice: on the fitted intensity
matrices $M_t$ and on their unit-mean versions
$S_t=M_t/\overline{M_t}$. The raw comparison retains total scale; the shape
comparison removes average intensity and focuses on redistribution.

The normalized Frobenius distance is the root-mean-square aligned cell
difference,

$$
d_F(M_1,M_2)=\frac{\lVert M_1-M_2\rVert_F}{n}.
$$

For $D=M_1-M_2$, the exact normalized node-aligned cut norm is

$$
d_{\square}^{\mathrm{aligned}}(M_1,M_2)=
\frac{1}{n^2}\max_{S,T\subseteq\{1,\ldots,n\}}
\left|\sum_{i\in S,\,j\in T}D_{ij}\right|.
$$

Exact computation is NP-hard. The implementation therefore reports
$\widehat d_{\square}^{\mathrm{aligned}}$, the best reproducible lower bound
found by alternating maximization on both $D$ and $-D$, starting from three
deterministic row subsets and 200 random restarts per sign in publication mode.
The best value among 1,000 random subset-pair draws is retained as a floor.
This quantity must be described as a **cut-norm lower-bound estimate**, not as
the exact cut norm and not as an unlabeled graphon cut distance.

The empirical one-dimensional Wasserstein distance is

$$
d_{W_1}(M_1,M_2)=\frac{1}{n^2}\sum_{r=1}^{n^2}
\left|x_{(r)}-y_{(r)}\right|,
$$

where $x_{(r)}$ and $y_{(r)}$ are sorted fitted entries. It compares intensity
distributions and deliberately discards origin--destination locations [5,6].
Finally, the original adjacency matrices are compared through the root-mean-
square difference between their ten leading singular values. This benchmark
captures dominant directed modes without relying on possibly complex
eigenvalues of a nonsymmetric matrix.

## Step 5. Quantify role instability and robustness

Structural-role change is measured as one minus the adjusted Rand index between
consecutive fitted partitions [7], so arbitrary block labels do not affect the
score. Estimator and resolution robustness are summarized by Spearman
correlation of all 18 transition scores and by Jaccard overlap between the two
top-five transition sets. Correlation measures overall rank stability; Jaccard
overlap tests whether the substantively highlighted transitions remain the
same.

## Step 6. Construct anomaly profiles and use them as a decision guide

For the empirical sequence, every transition score is converted to its
within-channel percentile among the 18 consecutive-year transitions. Correlated
channels are then grouped into six perspectives: activity and connectivity,
distribution and reciprocity, adjacency spectrum, raw fitted kernel, normalized
shape, and structural roles. The mean of the six perspective percentiles is a
consensus navigation score, and breadth is the number of perspectives at or
above the 75th percentile. Equal weighting at the perspective level prevents a
large family of closely related distances from dominating the summary merely
because it contains more channels. The 75th-percentile boundary is a descriptive
navigation convention, not a significance threshold; the full six-perspective
profile remains the primary empirical result. These quantities describe
relative anomaly within the observed period; they are not posterior
probabilities, formal change-point tests, or causal evidence.

No channel should be chosen only because it produces the largest standardized
peak. The research question determines the primary evidence, and at least one
channel from the other evidence family, with a different blind spot, should be
used as a complement.

| Research question | Primary evidence | What a large value supports | Main blind spot | Recommended complement |
|---|---|---|---|---|
| Did overall activity or connectivity change? | Total weight and edge density; raw distances | A change in aggregate scale or occupied dyads | Cannot separate broad scaling from structured redistribution | Compare raw and unit-mean shape channels |
| Was change widespread across aligned pairs? | Frobenius distance | Many moderate cell changes or a few large cell changes | Does not identify a coherent affected subset | Cut-norm lower-bound estimate and conventional indicators |
| Did a coordinated origin--destination subset change? | Cut-norm lower-bound estimate | A strong rectangular signed discrepancy | Lower bound is optimization- and resolution-sensitive | Frobenius distance plus restart and resolution checks |
| Did the distribution of fitted weights change? | Empirical $W_1$ distance | A shift in the marginal intensity distribution | Discards node-pair locations | Node-aligned Frobenius or cut-norm lower-bound estimate |
| Did dominant directed modes change? | Singular-spectrum distance | A change in leading adjacency-matrix modes | Does not localize nodes or block pairs | Kernel distances and indicators |
| Did mesoscale roles reorganize? | One minus adjusted Rand index | A label-invariant partition change | Estimated roles can be noisy and low-powered | Kernel distances plus role-flow interpretation |
| Is a reported transition robust? | Spearman rank agreement and top-five Jaccard overlap | Timing is not tied to one estimator or resolution | Agreement does not validate substantive meaning | Known-truth calibration and domain interpretation |

## Step 7. Calibrate channel sensitivity with known-truth simulations

The known-truth scenarios in Figure 2 isolate six conceptually different
structural interventions. Each simulation uses $n=80$ nodes and four latent
roles. Edges are sampled from a block-specific Bernoulli presence model and
positive weights from block-specific conditional means with node-level sender
and receiver heterogeneity. Publication mode uses 200 replicates for the null
and for each small, medium, and large non-null effect. Figure 2 reports the
relative change in the expected kernel at the medium effect level before and
after unit-mean normalization. It therefore shows, on an interpretable common
scale, whether a scenario primarily changes network magnitude or structural
shape.

```{r figure-simulation-scenarios, fig.width=6.8, fig.height=4.15, fig.cap="Relative change in the expected weighted kernel under six medium-effect simulation scenarios. Connected points compare raw intensity with unit-mean structural shape; labels report the actual relative Frobenius change rather than a column-normalized color scale."}
kernel_design <- utils::read.csv(figure_inputs["simulation_kernels"])

scenario_order <- c(
  "global_scale", "density_shift", "destination_attraction",
  "directional_shift", "local_block_shock", "role_reassignment"
)
scenario_labels <- kernel_design %>%
  dplyr::distinct(.data$scenario, .data$scenario_label)
scenario_labels <- stats::setNames(
  scenario_labels$scenario_label, scenario_labels$scenario
)

kernel_matrix <- function(data) {
  as.matrix(stats::xtabs(
    expected_weight ~ origin_block + destination_block, data = data
  ))
}
relative_kernel_change <- lapply(scenario_order, function(scenario_id) {
  scenario_data <- kernel_design[kernel_design$scenario == scenario_id, ]
  baseline <- kernel_matrix(scenario_data[scenario_data$state == "Baseline", ])
  changed <- kernel_matrix(scenario_data[scenario_data$state == "Changed", ])
  baseline_shape <- baseline / mean(baseline)
  changed_shape <- changed / mean(changed)
  data.frame(
    scenario = scenario_id,
    raw_change = sqrt(sum((changed - baseline)^2)) / sqrt(sum(baseline^2)),
    shape_change = sqrt(sum((changed_shape - baseline_shape)^2)) /
      sqrt(sum(baseline_shape^2))
  )
}) %>%
  dplyr::bind_rows()

scenario_response <- relative_kernel_change %>%
  dplyr::mutate(
    scenario_label = factor(
      unname(scenario_labels[.data$scenario]),
      levels = rev(unname(scenario_labels[scenario_order]))
    ),
    y_base = as.numeric(.data$scenario_label)
  )

scenario_points <- scenario_response %>%
  tidyr::pivot_longer(
    cols = c("raw_change", "shape_change"),
    names_to = "response", values_to = "relative_change"
  ) %>%
  dplyr::mutate(
    response = dplyr::recode(
      .data$response,
      raw_change = "Raw intensity",
      shape_change = "Unit-mean shape"
    ),
    y = .data$y_base + ifelse(.data$response == "Raw intensity", 0.12, -0.12),
    label = scales::percent(.data$relative_change, accuracy = 0.1)
  )

p_scenarios <- ggplot2::ggplot(
  scenario_response,
  ggplot2::aes(y = y_base)
) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = shape_change, xend = raw_change,
      yend = y_base
    ),
    color = mx_colors["mid_grey"], size = 1.0
  ) +
  ggplot2::geom_point(
    data = scenario_points,
    ggplot2::aes(x = relative_change, y = y, fill = response),
    shape = 21, size = 4.2, color = mx_colors["ink"], stroke = 0.35
  ) +
  ggplot2::geom_text(
    data = scenario_points,
    ggplot2::aes(x = relative_change, y = y, label = label),
    nudge_x = 0.015, hjust = 0, size = 3.0, color = mx_colors["ink"]
  ) +
  ggplot2::scale_fill_manual(
    values = c(
      "Raw intensity" = mx_colors["blue"],
      "Unit-mean shape" = mx_colors["gold"]
    ),
    name = "Comparison scale"
  ) +
  ggplot2::scale_x_continuous(
    limits = c(0, 0.75), breaks = seq(0, 0.7, by = 0.1),
    labels = scales::label_percent(accuracy = 1)
  ) +
  ggplot2::scale_y_continuous(
    breaks = seq_along(levels(scenario_response$scenario_label)),
    labels = levels(scenario_response$scenario_label),
    expand = ggplot2::expansion(add = 0.45)
  ) +
  ggplot2::labs(
    title = "Normalization reveals whether a perturbation changes magnitude or shape",
    subtitle = "Medium-effect expected-kernel response; connected points belong to the same scenario",
    x = "Relative Frobenius change in the expected kernel", y = NULL
  ) +
  theme_methodsx(9) +
  ggplot2::theme(
    panel.grid.major.y = ggplot2::element_blank(),
    legend.position = "bottom"
  )
save_methodsx_figure(p_scenarios, "FIG02_simulation_scenarios", 9.5, 5.8)
p_scenarios
```

The paired responses make the scale--shape distinction explicit. At the medium
effect, global volume scaling changes the raw expected kernel by 59.5% but its
unit-mean shape by only 2.5%; the density intervention changes the raw kernel by
45.0% and leaves the normalized expected shape unchanged. Destination
attraction, directional change, local block shock, and role reassignment retain
substantial shape responses. Raw and normalized comparisons therefore provide
complementary evidence rather than interchangeable versions of one distance.

## Reproducibility implementation

The complete workflow is executed by `source("run_all.R")`. Publication mode
uses random seed 20260817, 200 simulation replicates per design cell, 1,000
random cut-search floor samples for empirical comparisons, 128 for simulation
comparisons, and 200 alternating-maximization restarts. Diagnostic mode uses
smaller values for rapid error detection but writes the same file structure.
Every analytical table is checked before rendering. The final output manifest
is written only after the HTML manuscript, Word manuscript, and all PDF/PNG
figures have been generated, so its sizes and MD5 hashes refer to the current
run rather than to previously rendered files.

OpenAI ChatGPT was used in a supportive role for code review, code editing,
visualization redesign, and manuscript organization. It did not generate or
alter the empirical observations or simulation outcomes. Zsolt T. Kosztyán
reviewed the AI-assisted changes and verified the reported claims against the
scripted outputs. Kornél Dénes will independently validate the software and
computational claims before submission. The authors retain full responsibility
for the analytical code and manuscript.

# Method validation

## Known-truth simulation calibration

Figure 3 evaluates sensitivity after calibrating every channel against its own
null distribution. Each cell is the proportion of simulated transitions that
exceeded the channel-specific 95th-percentile null threshold. The null column
therefore checks false-positive control, while the remaining columns reveal
which structural changes each evidence channel is capable of detecting.

```{r figure-simulation-power, fig.width=6.8, fig.height=5.10, fig.cap="Null-calibrated detection rates by change scenario and evidence channel. Non-null columns show the medium effect level; the null column shows the empirical false-positive rate."}
simulation_power <- utils::read.csv(figure_inputs["simulation_power"])
selected_channels <- c(
  "indicator_total_weight", "indicator_edge_density",
  "indicator_weighted_reciprocity", "indicator_in_strength_gini",
  "indicator_effective_destinations", "graphon_frobenius_raw",
  "graphon_cut_raw", "graphon_frobenius_shape", "graphon_cut_shape",
  "graphon_wasserstein_shape", "singular_spectrum_raw",
  "estimated_role_instability"
)
null_power <- simulation_power %>%
  dplyr::filter(.data$scenario == "null", .data$channel %in% selected_channels) %>%
  dplyr::mutate(display_scenario = "Null false-positive rate")
medium_power <- simulation_power %>%
  dplyr::filter(
    .data$scenario != "null", .data$effect_label == "medium",
    .data$channel %in% selected_channels
  ) %>%
  dplyr::mutate(display_scenario = .data$scenario_label)
power_plot_data <- dplyr::bind_rows(null_power, medium_power)
scenario_order <- c("Null false-positive rate", unique(medium_power$display_scenario))
power_plot_data$display_scenario <- factor(
  power_plot_data$display_scenario, levels = scenario_order
)
power_plot_data$channel_label <- factor(
  unname(channel_labels[power_plot_data$channel]),
  levels = rev(unname(channel_labels[selected_channels]))
)
p_power <- ggplot2::ggplot(
  power_plot_data,
  ggplot2::aes(
    x = display_scenario, y = channel_label,
    fill = calibrated_detection_rate
  )
) +
  ggplot2::geom_tile(color = "white", size = 0.35) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = scales::percent(calibrated_detection_rate, accuracy = 1),
      color = calibrated_detection_rate > 0.55
    ),
    size = 2.8
  ) +
  ggplot2::scale_color_manual(values = c(`TRUE` = "white", `FALSE` = mx_colors["ink"]), guide = "none") +
  ggplot2::scale_fill_gradientn(
    colors = c("#F4F6F7", "#C9DCE5", mx_colors["blue"]),
    limits = c(0, 1), labels = scales::label_percent(), name = "Detection\nrate"
  ) +
  ggplot2::labs(
    title = "Different structural changes activate different evidence channels",
    subtitle = "Threshold for each channel: 95th percentile of its null distribution",
    x = NULL, y = NULL
  ) +
  theme_methodsx(9) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
    panel.grid = ggplot2::element_blank(), legend.position = "right"
  )
save_methodsx_figure(p_power, "FIG03_simulation_power", 10, 7.5)
p_power
```

The largest null detection rate is 5%, confirming the intended calibration in
this finite simulation. Conventional network indicators provide strong and
readily interpretable detection where the intervention targets their own
structural dimension. At the medium effect, total weight detects every global
scaling, density-shift, and destination-attraction replicate; edge density and
weighted reciprocity detect every density shift; and both in-strength Gini and
effective destinations detect every destination-attraction replicate.

Graphon-inspired and adjacency-matrix distances add evidence for changes that
are less completely summarized by annual indicators. Global scaling is
detected in all replicates by raw Frobenius, the raw cut-norm lower-bound
estimate, and the raw singular spectrum, while the corresponding unit-mean
shape distances remain near their null rates. Directional change is detected by
the raw singular spectrum in 88.0% of replicates and by the raw cut-norm lower-
bound estimate in 86.0%, compared with 71.5% for total weight and 57.5% for raw
Frobenius. For localized block shocks and role reassignment, the raw cut-norm
lower-bound estimate reaches 61.5% and 59.5%, compared with 57.0% and 37.0% for
total weight; estimated role instability reaches only 21.0% and 12.0%.

The result is complementarity rather than a universal winner. Conventional
indicators locate interpretable changes in network-level trajectories; graphon-
inspired distances test whether aligned relational patterns change beyond what
those summaries reveal; and the anomaly profile records whether the two
evidence families converge or disagree.

## Empirical anomaly profiles

The empirical illustration in Figure 4 asks a different question: which annual
transitions are relatively unusual across several structural perspectives?
Unlike the simulation rates, these are descriptive within-channel percentiles,
not null-hypothesis rejection probabilities. The first two perspectives retain
the conventional annual indicators; the adjacency spectrum, raw fitted kernel,
normalized shape, and structural roles add increasingly relational evidence.
Transitions are ordered by their mean perspective percentile, and the consensus
cell reports how many perspectives place the transition at or above the 75th
percentile.

```{r figure-empirical-evidence, fig.width=6.8, fig.height=4.29, fig.cap="Convergence of empirical transition evidence across six structural perspectives. The eight highest-consensus annual transitions are shown. The consensus cell reports the mean perspective percentile and, below it, the number of perspectives at or above the 75th percentile."}
empirical_evidence <- utils::read.csv(figure_inputs["empirical_evidence"])
evidence_channels <- c(
  "indicator_total_weight", "indicator_edge_density",
  "indicator_weighted_reciprocity", "indicator_in_strength_gini",
  "singular_spectrum_raw", "graphon_frobenius_raw", "graphon_cut_raw",
  "graphon_frobenius_shape", "graphon_cut_shape",
  "estimated_role_instability"
)
evidence_plot_data <- dplyr::bind_rows(
  empirical_evidence %>%
    dplyr::filter(
      .data$analysis_dimension %in% c("conventional", "adjacency benchmark"),
      .data$channel %in% evidence_channels
    ),
  empirical_evidence %>%
    dplyr::filter(
      .data$analysis_dimension == "estimator", .data$estimator == "block",
      .data$fixed_K == 13, .data$channel %in% evidence_channels
    )
)
evidence_plot_data$channel_label <- factor(
  unname(channel_labels[evidence_plot_data$channel]),
  levels = unname(channel_labels[evidence_channels])
)

perspective_lookup <- c(
  indicator_total_weight = "Activity and connectivity",
  indicator_edge_density = "Activity and connectivity",
  indicator_weighted_reciprocity = "Distribution and reciprocity",
  indicator_in_strength_gini = "Distribution and reciprocity",
  singular_spectrum_raw = "Adjacency spectrum",
  graphon_frobenius_raw = "Raw fitted kernel",
  graphon_cut_raw = "Raw fitted kernel",
  graphon_frobenius_shape = "Normalized shape",
  graphon_cut_shape = "Normalized shape",
  estimated_role_instability = "Structural roles"
)
perspective_order <- c(
  "Activity and connectivity", "Distribution and reciprocity",
  "Adjacency spectrum", "Raw fitted kernel", "Normalized shape",
  "Structural roles"
)
perspective_evidence <- evidence_plot_data %>%
  dplyr::mutate(
    perspective = unname(perspective_lookup[.data$channel])
  ) %>%
  dplyr::group_by(.data$transition, .data$perspective) %>%
  dplyr::summarise(
    perspective_percentile = mean(.data$evidence_percentile),
    .groups = "drop"
  )

transition_consensus <- perspective_evidence %>%
  dplyr::group_by(.data$transition) %>%
  dplyr::summarise(
    consensus = mean(.data$perspective_percentile),
    breadth = sum(.data$perspective_percentile >= 0.75),
    .groups = "drop"
  ) %>%
  dplyr::arrange(dplyr::desc(.data$consensus), dplyr::desc(.data$breadth)) %>%
  dplyr::mutate(consensus_rank = dplyr::row_number())

top_transitions <- utils::head(transition_consensus$transition, 8)
perspective_cells <- perspective_evidence %>%
  dplyr::filter(.data$transition %in% top_transitions) %>%
  dplyr::transmute(
    transition = .data$transition,
    perspective = .data$perspective,
    value = .data$perspective_percentile,
    label = scales::percent(.data$perspective_percentile, accuracy = 1)
  )
consensus_cells <- transition_consensus %>%
  dplyr::filter(.data$transition %in% top_transitions) %>%
  dplyr::transmute(
    transition = .data$transition,
    perspective = "Consensus",
    value = .data$consensus,
    label = paste0(
      "#", .data$consensus_rank, "  ",
      scales::percent(.data$consensus, accuracy = 1), "\n",
      .data$breadth, "/6"
    )
  )
consensus_plot_data <- dplyr::bind_rows(
  perspective_cells, consensus_cells
) %>%
  dplyr::mutate(
    transition = factor(.data$transition, levels = rev(top_transitions)),
    perspective = factor(
      .data$perspective, levels = c(perspective_order, "Consensus")
    )
  )

p_evidence <- ggplot2::ggplot(
  consensus_plot_data,
  ggplot2::aes(x = perspective, y = transition, fill = value)
) +
  ggplot2::geom_tile(color = "white", size = 0.45) +
  ggplot2::geom_text(
    ggplot2::aes(label = label, color = value >= 0.62),
    size = 2.8, lineheight = 0.9
  ) +
  ggplot2::geom_vline(xintercept = 6.5, color = mx_colors["ink"], size = 0.5) +
  ggplot2::scale_color_manual(
    values = c(`TRUE` = "white", `FALSE` = mx_colors["ink"]), guide = "none"
  ) +
  ggplot2::scale_fill_gradientn(
    colors = c("#F4F6F7", "#C9DCE5", mx_colors["blue"]),
    limits = c(0, 1), labels = scales::label_percent(), name = "Within-channel\npercentile"
  ) +
  ggplot2::labs(
    title = "The strongest transitions combine different structural signatures",
    subtitle = "Perspective-level evidence for the eight highest-consensus annual transitions",
    x = NULL, y = NULL
  ) +
  theme_methodsx(9) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
    panel.grid = ggplot2::element_blank(), legend.position = "right"
  )
save_methodsx_figure(p_evidence, "FIG04_empirical_transition_evidence", 9.5, 6)
p_evidence
```

The highest consensus occurs in 2017--2018 (mean perspective percentile 84.8%;
five of six perspectives at or above the 75th percentile). The next three are
2020--2021 (77.0%; three of six), 2011--2012 (70.6%; two of six), and
2022--2023 (69.1%; four of six). Their differing column profiles show why the
consensus score is a navigation aid rather than a replacement for the component
evidence. In 2017--2018, activity and connectivity are high (82.4%), but
distribution and reciprocity are not (35.3%); raw kernel, normalized-shape,
spectral, and role perspectives all exceed 94%. In 2022--2023, both conventional
indicator perspectives and the raw-kernel and spectral perspectives are high,
whereas normalized shape (50.0%) and roles (17.6%) are not. These profiles show
how graphon distances refine the interpretation of annual indicator changes
without displacing them. Importantly, 2015--2016 ranks only eleventh by
consensus and is not among the eight transitions displayed. The corrected
empirical input therefore does not support treating the transition adjacent to
2016 as a dominant break.

## Estimator and resolution robustness

Figure 5 examines whether the empirical ranking of annual transitions depends
on the fitted-kernel estimator or the selected fixed resolution. Cell color
reports Spearman rank correlation, while the annotation reports both that
correlation and the top-five Jaccard overlap. The two measures distinguish
overall rank stability from agreement about the most prominent transitions.

```{r figure-robustness, fig.width=6.8, fig.height=4.44, fig.cap="Agreement of empirical transition rankings across fitted-kernel estimators and fixed resolutions. Cell color and the first annotation line show Spearman correlation; the second line shows top-five Jaccard overlap."}
estimator_agreement <- utils::read.csv(figure_inputs["estimator_agreement"])
resolution_agreement <- utils::read.csv(figure_inputs["resolution_agreement"])
agreement_channels <- c(
  "graphon_frobenius_raw", "graphon_cut_raw",
  "graphon_frobenius_shape", "graphon_cut_shape"
)
estimator_heatmap <- estimator_agreement %>%
  dplyr::filter(.data$channel %in% agreement_channels) %>%
  dplyr::mutate(
    comparison_1 = as.character(.data$comparison_1),
    comparison_2 = as.character(.data$comparison_2),
    comparison = paste(.data$comparison_1, "vs", .data$comparison_2),
    panel = "Estimator agreement at K=13"
  )
resolution_heatmap <- resolution_agreement %>%
  dplyr::mutate(
    comparison_1 = as.character(.data$comparison_1),
    comparison_2 = as.character(.data$comparison_2)
  ) %>%
  dplyr::filter(
    .data$channel %in% agreement_channels,
    .data$comparison_1 != .data$comparison_2
  ) %>%
  dplyr::mutate(
    comparison = paste0("K=", .data$comparison_2, " vs K=", .data$comparison_1),
    panel = "Resolution agreement versus K=13"
  )
agreement_plot_data <- dplyr::bind_rows(
  estimator_heatmap, resolution_heatmap
)
agreement_plot_data$channel_label <- factor(
  unname(channel_labels[agreement_plot_data$channel]),
  levels = unname(channel_labels[agreement_channels])
)
agreement_plot_data$comparison <- factor(
  agreement_plot_data$comparison,
  levels = rev(unique(agreement_plot_data$comparison))
)
agreement_plot_data$panel <- factor(
  agreement_plot_data$panel,
  levels = c("Estimator agreement at K=13", "Resolution agreement versus K=13")
)
agreement_plot_data <- agreement_plot_data %>%
  dplyr::mutate(
    annotation = paste0(
      "rho ", sprintf("%.2f", .data$spearman_rho), "\n",
      "J ", scales::percent(.data$top5_jaccard, accuracy = 1)
    )
  )
p_agreement <- ggplot2::ggplot(
  agreement_plot_data,
  ggplot2::aes(
    x = channel_label, y = comparison, fill = spearman_rho
  )
) +
  ggplot2::geom_tile(color = "white", size = 0.5) +
  ggplot2::geom_text(
    ggplot2::aes(label = annotation, color = abs(spearman_rho) >= 0.68),
    size = 2.8, lineheight = 0.9
  ) +
  ggplot2::facet_grid(panel ~ ., scales = "free_y", space = "free_y") +
  ggplot2::scale_color_manual(
    values = c(`TRUE` = "white", `FALSE` = mx_colors["ink"]), guide = "none"
  ) +
  ggplot2::scale_fill_gradient2(
    low = mx_colors["orange"], mid = "#F4F6F7", high = mx_colors["blue"],
    midpoint = 0, limits = c(-1, 1), name = "Spearman\ncorrelation"
  ) +
  ggplot2::labs(
    title = "Robustness differs by distance and specification choice",
    subtitle = "rho = Spearman correlation; J = top-five Jaccard overlap",
    x = NULL, y = NULL
  ) +
  theme_methodsx(9) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1),
    panel.grid = ggplot2::element_blank(), legend.position = "right"
  )
save_methodsx_figure(p_agreement, "FIG05_robustness_agreement", 9.5, 6.2)
p_agreement
```

Raw-scale Frobenius is the most stable displayed channel: estimator correlations are
0.876--0.963, and correlations with the $K=13$ ranking are 0.756, 0.957, and
0.938 at $K=2,4,$ and 8. The raw-scale cut-norm lower-bound estimate is robust
across estimators (0.765--0.899) but more resolution-sensitive (0.542, 0.870,
and 0.676). The unit-mean shape cut-norm lower-bound estimate is the clearest
warning case: its $K=8$ correlation with $K=13$ is
-0.003 and its top-five overlap is 11.1%. The diverging color scale is necessary
to display this near-zero negative association rather than suppress it. A
transition supported only by the shape-scale cut estimate should therefore be
treated as specification-sensitive; convergence across raw, shape, spectral,
and indicator perspectives supports a stronger empirical claim.

## Structural interpretation through role flows

Figure 6 moves from detecting a transition to interpreting a selected network
state. A single role-flow map encodes two non-equivalent quantities: bubble area
is a role pair's share of all observed application weight, whereas fill is the
share conditional on its origin role. The design therefore makes systemwide
concentration and origin-specific destination choice visible without using
separately normalized heatmaps.

```{r figure-role-flow, fig.width=6.8, fig.height=5.60, fig.cap="Observed application-weight flow among active fixed-K=13 destination-oriented structural roles in 2018. Bubble area represents each role pair's share of total system weight; fill represents its share conditional on the origin role. Role rank 1 is the strongest fitted destination role."}
role_flow <- utils::read.csv(figure_inputs["role_flow"])
active_origins <- role_flow %>%
  dplyr::group_by(.data$origin_role_rank) %>%
  dplyr::summarise(total = sum(.data$observed_weight), .groups = "drop") %>%
  dplyr::filter(.data$total > 0) %>%
  dplyr::pull(.data$origin_role_rank)
active_destinations <- role_flow %>%
  dplyr::group_by(.data$destination_role_rank) %>%
  dplyr::summarise(total = sum(.data$observed_weight), .groups = "drop") %>%
  dplyr::filter(.data$total > 0) %>%
  dplyr::pull(.data$destination_role_rank)

role_flow_active <- role_flow %>%
  dplyr::filter(
    .data$origin_role_rank %in% active_origins,
    .data$destination_role_rank %in% active_destinations
  ) %>%
  dplyr::group_by(.data$origin_role_rank) %>%
  dplyr::mutate(
    origin_conditional_share = .data$observed_weight / sum(.data$observed_weight)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    label = ifelse(
      .data$observed_weight_share >= 0.015,
      scales::percent(.data$observed_weight_share, accuracy = 0.1),
      ""
    ),
    light_text = .data$origin_conditional_share >= 0.45 & .data$label != ""
  )

destination_marginals <- role_flow %>%
  dplyr::group_by(.data$destination_role_rank) %>%
  dplyr::summarise(share = sum(.data$observed_weight_share), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(.data$share))
top_destination_share <- sum(utils::head(destination_marginals$share, 3))
top_four_flow_share <- sum(
  utils::head(sort(role_flow$observed_weight_share, decreasing = TRUE), 4)
)

p_role_flow <- ggplot2::ggplot(
  role_flow_active,
  ggplot2::aes(
    x = destination_role_rank, y = origin_role_rank,
    size = observed_weight_share, fill = origin_conditional_share
  )
) +
  ggplot2::geom_point(
    shape = 21, color = mx_colors["ink"], stroke = 0.25, alpha = 0.95
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = label, color = light_text),
    size = 2.4, fontface = "bold", show.legend = FALSE
  ) +
  ggplot2::scale_color_manual(
    values = c(`TRUE` = "white", `FALSE` = mx_colors["ink"]), guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    breaks = active_destinations,
    expand = ggplot2::expansion(add = 0.6)
  ) +
  ggplot2::scale_y_reverse(
    breaks = active_origins,
    expand = ggplot2::expansion(add = 0.6)
  ) +
  ggplot2::scale_fill_gradient(
    low = "#F3F6F8", high = mx_colors["orange"], limits = c(0, 1),
    labels = scales::label_percent(accuracy = 1),
    name = "Share within\norigin role"
  ) +
  ggplot2::scale_size_area(
    max_size = 13,
    labels = scales::label_percent(accuracy = 0.1),
    name = "Share of all\napplication weight"
  ) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Role flows separate systemwide concentration from origin-specific choice",
    subtitle = "Bubble area = total-system share; fill = conditional destination share",
    x = "Destination role rank", y = "Origin role rank"
  ) +
  theme_methodsx(9) +
  ggplot2::theme(
    legend.position = "right",
    panel.grid.minor = ggplot2::element_blank()
  )
save_methodsx_figure(p_role_flow, "FIG06_role_flow_matrix", 8.5, 7)
p_role_flow
```

The flow structure is strongly concentrated: the three largest destination
roles receive `r scales::percent(top_destination_share, accuracy = 0.1)` of all
application weight, and the four largest origin--destination role pairs account
for `r scales::percent(top_four_flow_share, accuracy = 0.1)`. The conditional
fill nevertheless reveals differences among origin roles that bubble area alone
would obscure. This illustrates why role-flow interpretation adds substantive
information after a structural transition has been identified; it does not
serve as an independent change detector.

# Practical interpretation and conclusions

The workflow should be read as a three-part evidence-integration architecture. First,
conventional annual network indicators preserve substantive interpretability
and show how activity, connectivity, reciprocity, and concentration evolve.
Second, graphon-inspired and adjacency-based distances test whether the aligned
relational pattern changes in ways that scalar summaries can miss. Third, the
anomaly-scoring, simulation-calibration, and consensus layer makes these
heterogeneous signals comparable and records the breadth of their agreement.
The third part is an integration and validation layer, not another source of
network data.

Accordingly, reporting should begin with the annual network-indicator
trajectories, then use graphon-inspired distances to refine the type and
location-scale of structural change, and finally use anomaly profiles,
specification checks, and role flows to assess robustness and interpretation.
This ordering prevents a graphon-only reading of the method while also avoiding
the opposite mistake of treating annual scalar indicators as a complete
description of network evolution.

The known-truth results reject a universal ranking of methods: indicator,
kernel-distance, spectral, and role channels respond differently to the same
perturbation. The appropriate final claim is therefore conditional. Convergent
indicator and graphon-inspired evidence supports a broad structural anomaly;
divergence identifies whether the signal is mainly aggregate, relational,
scale-dependent, shape-dependent, spectral, or role-based. Researchers should
report that profile rather than average away the disagreement.

# Limitations

The workflow assumes that node identity and weight meaning are comparable over
time. If nodes enter or leave the system, researchers must first define a common
universe or use a method designed for unmatched networks. Because the reported
distances are node-aligned, they are not invariant to relabeling and should not
be described as unlabeled graphon distances.

The cut quantity is a reproducible lower-bound estimate. Alternating
maximization can converge to a local optimum, and increasing the number of
restarts may increase the reported score. Absolute cut values should therefore
be accompanied by the restart count and random-search floor; transition
rankings should be checked across resolutions.

Block-kernel results depend on estimator and resolution. A fixed $K$ makes
annual distances comparable but does not establish that $K$ is the true latent
number of roles. The empirical robustness analysis shows that raw Frobenius is
comparatively stable in this application, whereas shape cut can be highly
resolution-sensitive. These findings are data-specific and do not establish a
universal ranking of distance measures.

The simulations isolate seven deliberately simplified cases, including the
null. They evaluate detection under the documented $n=80$, $K=4$ weighted block
model and effect schedule; they do not prove sensitivity under every sparse,
heavy-tailed, dependent, or node-misaligned process. Researchers should add
domain-relevant perturbations when applying the workflow elsewhere.

Empirical within-channel percentiles are descriptive relative anomaly scores
over the 18 observed transitions. They are not calibrated $p$-values, do not
correct for multiple testing, and do not by themselves establish a formal
change point. The equal perspective weights, perspective definitions, and
75th-percentile breadth boundary are transparent analytic choices rather than
estimated optimal parameters. Confirmatory use requires an external null model,
sensitivity analysis for alternative aggregation rules, or an appropriate
time-series change-point procedure.

Structural-role groups describe similarity of directed connection profiles;
they are not modularity communities. Role flows are descriptive aggregates and
do not identify individual pathways or causal mechanisms. In the empirical
application, edges represent applications rather than realized enrollment, so
destination-role position is not an institutional quality ranking and annual
associations are not causal effects of policy events.

# Ethics statements

The method development and reported illustration use aggregate annual
micro-region flow counts. The reproducibility package contains no applicant-
level records or direct personal identifiers, and no experimental intervention,
animal research, or new recruitment of human participants was undertaken for
this methods study. Access to and any further use of the underlying
administrative records remain subject to the data provider's authorization and
applicable data-protection requirements.

# CRediT author statement

Zsolt T. Kosztyán: Conceptualization; Methodology; Software; Validation; Formal
analysis; Investigation; Data curation; Visualization; Writing -- original
draft; Writing -- review and editing; Project administration.

Kornél Dénes: Software; Validation; Writing -- review and editing.

<!-- PRE-SUBMISSION AUTHORSHIP GATE: confirm that independent validation is
complete, both authors approve the final manuscript, and both accept
accountability for their contributions before removing draft status. -->

# Acknowledgments

The authors thank the colleagues who supported the reconstruction and checking
of the annual inputs. Project no. K 142395 was
implemented with support provided by the Ministry of Culture and Innovation of
Hungary from the National Research, Development and Innovation Fund, financed
under the K-22 "OTKA" funding scheme.

# Declaration of interests

The authors declare that they have no known competing financial interests or
personal relationships that could have appeared to influence the work reported
in this paper.

# Supplementary material and/or additional information [OPTIONAL]

The accompanying reproducibility package contains the R pipeline, aggregate
annual edge lists and node metadata, derived empirical and simulation tables,
validation reports, session information, output manifests, and publication-
quality PDF and PNG figures. Supplementary Figure S1 reports whether the fitted
role solutions are approximately nested across adjacent fixed resolutions.

```{r figure-supplement-refinement, fig.width=6.8, fig.height=4.68, fig.cap="Empirical consistency with coarse-to-fine role refinement across selected years and adjacent fixed resolutions."}
refinement <- utils::read.csv(figure_inputs["refinement"])
refinement$transition <- factor(
  paste0("K=", refinement$low_K, " to K=", refinement$high_K),
  levels = paste0(
    "K=", c(2, 4, 8), " to K=", c(4, 8, 13)
  )
)
refinement$year <- factor(refinement$year, levels = sort(unique(refinement$year)))
refinement$year_position <- as.numeric(refinement$year) +
  (as.numeric(refinement$transition) - 2) * 0.14
refinement_colors <- stats::setNames(
  unname(mx_colors[c("blue", "gold", "olive")]),
  levels(refinement$transition)
)
p_refinement <- ggplot2::ggplot(
  refinement,
  ggplot2::aes(
    x = weighted_refinement_consistency, y = year_position,
    fill = transition, shape = transition
  )
) +
  ggplot2::geom_vline(xintercept = 0.8, color = mx_colors["mid_grey"], linetype = 2) +
  ggplot2::geom_point(size = 4.2, color = mx_colors["ink"], stroke = 0.3) +
  ggplot2::scale_x_continuous(limits = c(0, 1), labels = scales::label_percent()) +
  ggplot2::scale_y_continuous(
    breaks = seq_along(levels(refinement$year)),
    labels = levels(refinement$year)
  ) +
  ggplot2::scale_fill_manual(
    values = refinement_colors,
    name = "Resolution step"
  ) +
  ggplot2::scale_shape_manual(
    values = stats::setNames(c(21, 22, 24), levels(refinement$transition)),
    name = "Resolution step"
  ) +
  ggplot2::labs(
    title = "Approximate refinement across fixed resolutions",
    subtitle = "Weighted share of nodes assigned consistently with dominant lower-resolution parents",
    x = "Weighted refinement consistency", y = NULL
  ) +
  theme_methodsx(9)
save_methodsx_figure(p_refinement, "FIGS01_refinement_consistency", 8, 5.5)
p_refinement
```

# Declaration of generative AI and AI-assisted technologies in the manuscript preparation process

During the preparation of this work, the authors used OpenAI ChatGPT to support
code review, code editing, visualization redesign, manuscript organization, and
English-language revision. After using this service, the authors reviewed and
edited the content, verified the computational claims against the reproducible
R outputs, and take full responsibility for the content of the published
article. No generative-AI system is listed as an author.

# References

[1] Donnat C, Holmes S. Tracking network dynamics: A survey using graph
distances. *Annals of Applied Statistics*. 2018;12(2):971--1012.
https://doi.org/10.1214/18-AOAS1176.

[2] Lovász L. *Large Networks and Graph Limits*. Providence, RI: American
Mathematical Society; 2012. https://doi.org/10.1090/coll/060.

[3] Wills P, Meyer FG. Metrics for graph comparison: A practitioner's guide.
*PLoS ONE*. 2020;15(2):e0228728.
https://doi.org/10.1371/journal.pone.0228728.

[4] Lovász L, Szegedy B. Limits of dense graph sequences. *Journal of
Combinatorial Theory, Series B*. 2006;96(6):933--957.
https://doi.org/10.1016/j.jctb.2006.05.002.

[5] Peyré G, Cuturi M. Computational optimal transport: With applications to
data science. *Foundations and Trends in Machine Learning*.
2019;11(5--6):355--607. https://doi.org/10.1561/2200000073.

[6] Villani C. *Optimal Transport: Old and New*. Berlin: Springer; 2009.
https://doi.org/10.1007/978-3-540-71050-9.

[7] Hubert L, Arabie P. Comparing partitions. *Journal of Classification*.
1985;2:193--218. https://doi.org/10.1007/BF01908075.

[8] White HC, Boorman SA, Breiger RL. Social structure from multiple networks.
I. Blockmodels of roles and positions. *American Journal of Sociology*.
1976;81(4):730--780. https://doi.org/10.1086/226141.
